\documentclass{article}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{mathtools}
\usepackage{blindtext}
\usepackage{multicol}
\usepackage{graphicx}
\usepackage{placeins}
\usepackage{algpseudocode}
\usepackage{algorithm}
\usepackage{booktabs}
\usepackage{float}
\usepackage{listings}
\usepackage{bbm}
\usepackage{wrapfig}
\usepackage{color}
\usepackage[bookmarks=true]{hyperref}
\usepackage{bookmark}
\graphicspath{ {./images/} }
\usepackage[bottom=0.5cm, right=1.5cm, left=1.5cm, top=1.5cm]{geometry}

\newtheorem{theorem}{Theorem}
\newtheorem{exercise}{Exercise}
\newtheorem{example}{Example}
\newtheorem{definition}{Definition}[section]

\newcommand{\R}{\ensuremath{\mathbb{R}}}
\newcommand{\N}{\ensuremath{\mathbb{N}}}
\newcommand{\Z}{\ensuremath{\mathbb{Z}}}
\newcommand{\p}{\ensuremath{\mathbb{P}}}
\newcommand{\E}{\ensuremath{\mathbb{E}}}
\newcommand{\F}{\ensuremath{\mathcal{F}}}
\newcommand{\1}{\ensuremath{\mathbbm{1}}}
\newcommand{\B}{\ensuremath{\mathbbm{B}}}

\definecolor{dkgreen}{rgb}{0,0.6,0}
\definecolor{gray}{rgb}{0.5,0.5,0.5}
\definecolor{mauve}{rgb}{0.58,0,0.82}

\lstset{frame=tb,
  language=C,
  aboveskip=3mm,
  belowskip=3mm,
  showstringspaces=false,
  columns=flexible,
  basicstyle={\small\ttfamily},
  numbers=none,
  numberstyle=\tiny\color{gray},
  keywordstyle=\color{blue},
  commentstyle=\color{dkgreen},
  stringstyle=\color{mauve},
  breaklines=true,
  breakatwhitespace=true,
  tabsize=3
}

\title{Lecture 12 - File systems}
\author{Gidon Rosalki}
\date{2025-06-22}

\begin{document}
\maketitle
\noindent\textbf{Notice:} If you find any mistakes, please open an issue at \href{https://github.com/robomarvin1501/notes\_operating\_systems}{\texttt{https://github.com/robomarvin1501/notes\_operating\_systems}}

\vspace{1cm}  % Add space before the rest of the content

\begingroup
\centering
{\color{red}\bfseries\LARGE This content will not appear in the exam this year! \par}
\endgroup

\vspace{1cm}  % Add space before the rest of the content

\section{Files}\label{sec:Files} % (fold)
A file is a named persistent sequential data storage. Sequential indicates that the data within the file is ordered, and
may have some structure. Persistent indicates that it lasts between restarts, and named means that it can be found based
off its name.

\subsection{Names}\label{sub:} % (fold)
Names are external to the operating system, and exist in many contexts: \begin{itemize}
    \item Access hsared memory segments (shmget, shmat)
    \item Domain name service to find web hosts
    \item Well known ports for certain services
\end{itemize}
One needs to remember the name to access the file, or possibly find it in a list, though this can take a long time. A
modern alternative is to search by content.
% subsection  (end)

\subsection{Persistence}\label{sub:Persistence} % (fold)
Files are stored on non-volatile devices, such as disks and other magnetic media, SSDs (non-volatile memory), and so
data is retained when computer turned off. Data also survives the process which created the data. Typically one process
creates the file, and others read the data later. This is as opposed to memory which is created and dies with the
process.
% subsection Persistence (end)

\subsection{Structure and meaning}\label{sub:Structure and meaning} % (fold)
In unix, data is a sequence of byutes. It can represent text, pictres, video, sound, program source code, program
binary, numerical data, and so on. In IBM mainframes, data is a sequence of records for databases. Finally, in windows
NTFS, it is a set of attribute value pairs, e.g. the file's icon with an indication of the internet source. It always
has an unnamed data stream wit contents.

\subsubsection{Filename extensions}\label{sec:Filename extensions} % (fold)
By \textbf{convention}, the filename is divided into 2 parts, the name, and the extension. However, extension-less files
are also common in Linux. Applications such as a C compiler might insist on extensions.
% subsubsection Filename extensions (end)
% subsection Structure and meaning (end)

\subsection{OS connection}\label{sub:OS connection} % (fold)
There needs to be a separation of concerns between the OS and the application.
Operating system: \begin{itemize}
    \item Provide infrastructure and support
    \item Store and retrieve the data
    \item Keep metadata about the data
    \item Don’t limit the application
\end{itemize}

Application: \begin{itemize}
    \item Do interpretation and processing
    \item Create and use the data
    \item Decide on data format (how things are represented)
    \item Know what the data means
\end{itemize}

The file system has an API, which abstracts files, directories and naming, and sharing and protection. To implement a
file system, the OS needs to be responsible for storing the data, space allocation methods and management, and provide
reliability and performance.

The FS operations are \textbf{system calls}. The OS implementation uses privileged instructions to activate storage
devices, and devices report completion using interrupts.

There are a few popular file systems. Linux generally uses ext4, the latest revision of the extended file system. The
previous version, ext3 could only support disks up to 16TB, which have since become more common, along with even larger
disks. We will mainly focus on ext3, but there is a lot of equivalence between it and ext4. Windows just uses NTFS, and
finally USB sticks tend to use FAT32 or exFAT. FAT32 can only support a limited size of files, so everyone is moving
over to exFAT (you are recommended to always reformat new USB sticks to exFAT, if they were originally FAT32).

\subsubsection{Basic file system operations}\label{sec:Basic file system operations} % (fold)
Have a nice big list of basic file system operations:
\begin{itemize}
    \item Create a file
    \item Delete a file
    \item Open a file
    \item Close a file
    \item Write to a file
    \item Read from a file
    \item Seek to somewhere in a file
    \item Truncate a file
    \item Set attributes of a file
    \item Get attributes of a file
    \item Rename a file
\end{itemize}
% subsubsection Basic file system operations (end)

\subsubsection{File attributes}\label{sec:File attributes} % (fold)
File specific info is maintained by the OS, called the metadata. This varies a lot across different OSs. There are two
parts to the metadata, the user facing, and the internal. \\
User facing: This contains things like the name (used to identify the file, human readable), the type (needed for
systems that support different types), the size, the owner user identification (for protection and security), access
rights (who can read, write, execute), and the time of creation, modification, and access. This last field is used
extensively by GNU make. \\
The internal OS use parts include the identifier, a unique tag (number) that identifies the file within the file system,
and the location, which is a pointer to the file location on the device.
% subsubsection File attributes (end)
% subsection OS connection (end)

% section Files (end)

\section{Files namespace}\label{sec:Files namespace} % (fold)
The motivation behind naming the file is to re-access the data that was stored previously, such that the user need not
remember the block, sector, and so on. This way a process can create a file, and give it a name, and the human can later
on find it in another process, and read it.

Naming conventions are OS dependent. Usually names as long as 255 characters are allowed, and digits and special
characters (such as space) are sometimes allowed. MS-DOS and Windows are not case sensitive (for they are fools), where
the UNIX family is. The naming administration is done separately to the data.

\subsection{Organisation}\label{sub:Organisation} % (fold)
Files are organised in a hierarchy of directories, which aim to obtain: \begin{enumerate}
    \item Efficiency - locating a file quickly, we want to avoid a single huge list with all the files
    \item Separation - convenient to users. This way two users can have the same name for different files, and the same
        file can have several different names
    \item Grouping: Logical grouping of files based off contents, properties, or usage. E.g. all java programs, all
        games, and the files related to OS exercise 4, and so on.
\end{enumerate}

\subsubsection{Directories}\label{sec:Directories} % (fold)
Directories map names to file objects. They are one level of the naming hierarchy. This data is stored in files, so
directories are in fact files themselves. A bit in the file metadata object (inode) indicates if it is a directory, and
the OS interprets the format of the directory file data. One should ideally use an efficient data structure, to enable
speed in searching / other file operations.

We can achieve this by having the namespace have a tree structure, of arbitrary height. Nodes are directories and files,
and internal nodes are directories. Directories contain files, and sub-directories. Names are relative to a directory,
and files are identified by their path from the root.

File systems have a root directory, and processes have a working directory. This is typically the home directory of
their user. To access a file, the process needs to specify the path in which the file is located. Path names are either
absolute or relative: \begin{itemize}
    \item  Absolute: path of file from the root directory
    \item  Relative: path from the current working directory
\end{itemize}
Path components are separated by / (unix) or \\ (windows, like fools). Most OSes have two special entries in each
directory: "." for current directory ".." for parent.

We can also have acyclic links between directories, where we have files located under two different directories
simultaneously (remember, directories are also files). This can be done through hard links and soft links.
\begin{itemize}
    \item Hard links: We can only perform these for files, not directories, to avoid recursive loops. They may also only
        be created across the same file system. They point at files (not names), and so are move safe. They are also
        reference counted, and so are delete safe
    \item Soft links (AKA symbolic links): These are links between any file and directory, and are neither move, nor
        delete safe. Should you delete a file, the soft link will still exist.
\end{itemize}
% subsubsection Directories (end)

\subsubsection{Protection}\label{sec:Protection} % (fold)
File owner / creator should be able to control what can be done, by whom. There are 6 types of access, Read, Write,
Execute, Append, (Dir) Delete, (Dir) List. Note, creating and deleting files are operations on the \textbf{directory}.

We have some categories of users to consider when it comes to protection and access rights. The individual user, who
establishes a user id on log in, which might be just local on the computer, or could be through interaction with a
network service, and groups, to which users belong, such as wheel, or docker.

To express access rights, we have two methods: \begin{itemize}
    \item Minimal (*nix): Just a single user, group, and the rest, which is compact enough to fit in just a few bytes.
        However, this is not very expressive.
    \item Detailed (Windows): A per file list that tells exactly who can do what to that file, which is useful in its
        expressiveness, but is in no way compact
\end{itemize}

Linux access rights dictate 3 modes of access: read, write, and execute. There are 3 classes of users, the owner, the
group, and the others. For each file or directory, we define desired access rights as 9 bits. Each group of 3 indicates
RWX for each of the groups.
% subsubsection Protection (end)
% subsection Organisation (end)
% section Files namespace (end)

\section{Layout and Access}\label{sec:Layout and Access} % (fold)
When accessing a file, we maintain our current location in an open file, and access is relative to this location. This
location can be changed based off \lstinline[columns=fixed]{rewind} or \lstinline[columns=fixed]{seek}.

How we layout the information on the disk can cause problems. It we simply store data contiguously, this can lead to
fragmentation, where there is enough memory to store a file, but it is not contiguous. There are a couple of possible
solutions: \begin{itemize}
    \item Compaction: Move around data in disk, to unify holes and create large enough holes to accommodate future
        files. This may improve performance thanks to serial access, however this is very costly, and there is not
        necessarily enough space to move around data like this.
    \item Divide files into fixed size blocks (usually 4KB), and all operations are made on blocks. This is similar to
        paging in memory management.
\end{itemize}

When using a block system, the user views a file as contiguous data, the API views it as a sequence of bytes, of
arbitrary size, the FS views it as a collection of 4KB blocks, and the disk views it as individual sectors (512 bytes).
Accesses are not necessarily aligned with blocks, and blocks may span multiple sectors.
% section Layout and Access (end)

Note, these notes are incomplete in comparison to the lecture. I ran out of energy, especially since the content is
extra (though fascinating). % TODO


\end{document}
